---
title: "CCLE LoY"
author: "Teagen Allen"
date: "2024-07-06"
output: html_document
---

Project Overview
This project analyzes gene expression data from the Cancer Cell Line Encyclopedia (CCLE) and copy number data for sex chromosomes from the paper "Gene Essentiatlity" by Shohat et al., 2022. The goal is to explore the relationship between gene expression and the ploidy of sex chromosomes.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Libraries
The libraries used in this analysis include `ggplot2` for plotting and `readr` for reading CSV files.

```{r libraries}
library(ggplot2)
library(dplyr)
library(UpSetR)

```

#### Directories
Set the working directory and data directory paths.

```{r directories}
# Set working and data directories
working_directory <- "C:/Users/teage/Documents/CCLE_LoY"  
data_directory <- "C:/Users/teage/Documents/CCLE_LoY/data/"
output_directory = "C:/Users/teage/Documents/CCLE_LoY/output/"

# Set working directory
setwd(working_directory)

```

#### Data Loading
Read the gene expression data and the supplemental table from the CSV files.

```{r data}
library(readr)

#chrY_gene_expression <- read_csv(data_directory/"chrY_gene_expression_annotated.csv")
Supplemental_Table_S1_cleaned <- read_csv("data/Supplemental_Table_S1_cleaned.csv")
CCLE_data <- read_delim(paste0(data_directory, "CCLE_RNAseq_rsem_genes_tpm_20180929.txt"), 
                          delim = "\t", escape_double = FALSE, trim_ws = TRUE)

annotation_data <- read.csv(paste0(data_directory,"Cell_lines_annotations_20181226.txt"), header = TRUE, sep = "\t")

```

```{r}
# Define the list of gene IDs and their corresponding gene names
gene_info <- c("ENSG00000067048.12" = "DDX3Y", 
               "ENSG00000198692.5"  = "EIF1AY", 
               "ENSG00000012817.11" = "KDM5D", 
               "ENSG00000165246.8"  = "NLGN4Y", 
               "ENSG00000129824.11" = "RPS4Y1", 
               "ENSG00000154620.5"  = "TMSB4Y", 
               "ENSG00000223555.1"  = "USP9Y", 
               "ENSG00000183878.11" = "UTY", 
               "ENSG00000067646.7"  = "ZFY", 
               "ENSG00000229807.5"  = "XIST")

# Subset the gene expression data by the chosen genes
chosen_genes_data <- subset(CCLE_data, gene_id %in% names(gene_info))

# Step to rename the gene IDs in the data frame
chosen_genes_data$gene_id <- gene_info[chosen_genes_data$gene_id]

# Remove the transcript_ids column
chosen_genes_data <- chosen_genes_data %>% select(-transcript_ids)

# Display the first few entries to verify the data
print(head(chosen_genes_data))


```

```{r}
# Transpose the data
transposed_data <- as.data.frame(t(chosen_genes_data))

# Set the first row as the header (gene names)
colnames(transposed_data) <- transposed_data[1, ]
transposed_data <- transposed_data[-1, ]

# Convert all columns to numeric
transposed_data[] <- lapply(transposed_data, as.numeric)

# Add a name to the first column
transposed_data <- cbind(sample = rownames(transposed_data), transposed_data)

# Reset row names to avoid duplication
rownames(transposed_data) <- NULL

# Display the first few entries of the transposed data
print(head(transposed_data))

# Now merge with the sex annotation data
sex_annotation <- annotation_data[, c("CCLE_ID", "Gender")]
colnames(sex_annotation)[1] <- "sample"  # Rename for merging
colnames(sex_annotation)[2] <- "reported_sex"  # Rename for accuate language

# Merge the sex annotation to the transposed data
transposed_data_with_sex <- merge(transposed_data, sex_annotation, by = "sample", all.x = TRUE)

# Display the first few entries to verify the merge
print(head(transposed_data_with_sex))


```

#### Data Merging
Merge the gene expression data with the supplemental table on the `cell_line` column after removing rows with `NA` values in the `cell_line` column.

```{r merge data}
# Subset the relevant columns
ploidy_data <- Supplemental_Table_S1_cleaned[, c("cell_line", "Biological_sex" , "Y_ploidy", "X_ploidy", "Genotype")]

# Rename DepMap_ID to match the sample column name in transposed_data
colnames(ploidy_data)[1] <- "sample"  # Rename for merging

# Merge with the transposed data
expression_ploidy_data <- merge(transposed_data, ploidy_data, by = "sample", all.x = TRUE)

# Display the first few entries to verify the merge
print(head(expression_ploidy_data))

```

#### Log Transformation
Perform log transformation on the expression columns to handle log(0) issues.
```{r}
# Define the gene columns (excluding 'sample' and any non-gene columns)
gene_columns <- colnames(expression_ploidy_data)[2:(ncol(expression_ploidy_data) - 4)]  # Adjust based on the number of non-gene columns

# Log-transform each gene expression value and create new columns
for (gene in gene_columns) {
  new_col_name <- paste0("log_", gene)  # Create new column name
  expression_ploidy_data[[new_col_name]] <- log10(expression_ploidy_data[[gene]] + 1)  # Log-transform and add to dataframe
}

# Display the first few entries to verify the log transformation
print(head(expression_ploidy_data))


```

```{r}
# Subset data into male and female
male_data <- expression_ploidy_data[expression_ploidy_data$Biological_sex == "Male", ]
female_data <- expression_ploidy_data[expression_ploidy_data$Biological_sex == "Female", ]

```


#### Plotting Gene Expression vs. Y Ploidy (NA Omitted)
Create function for violin plots for each gene expression variable grouped by Y ploidy, excluding rows with `NA` values.

```{r}
# Function to create violin plots for gene expression variables by sex in the same graph
create_gene_violin_plot_by_sex <- function(data, gene_expression, ploidy, remove_na = TRUE) {
  # Optionally filter out rows where ploidy is NA
  if (remove_na) {
    data <- data[!is.na(data[[ploidy]]), ]
  }

  # Filter out rows with "Miss-classification" in Biological_sex
  data <- data[data$Biological_sex != "Miss_classification", ]

  # Create an interaction variable for plotting
  data$interaction <- interaction(data$Biological_sex, data[[ploidy]], sep = "_")

  # Create spaced interaction by adding a space level
  data$spaced_interaction <- as.factor(paste(data$interaction, " ", sep = ""))

  ggplot(data, aes(x = spaced_interaction, y = !!sym(gene_expression), fill = Biological_sex)) + 
    geom_violin(width = 0.8, scale = "width", aes(fill = Biological_sex)) +  # Adjust width for better appearance
    labs(
      title = paste("Distribution of", gene_expression, "by Reported Sex and Ploidy"),
      x = "Reported Sex and Ploidy Status",
      y = paste("Log-transformed Expression of", gene_expression)
    ) +
    theme_minimal() +
    scale_y_continuous(labels = function(x) ifelse(is.finite(x), round(x, 2), NA)) +  # Handle NA values in y-axis labels
    scale_fill_manual(values = c("Female" = "darkblue", "Male" = "lightblue"), 
                      name = "Reported Sex") +
    scale_x_discrete(name = "Reported Sex and Ploidy Status")
}

```


```{r Plot Y Ploidy na.omit}
create_gene_violin_plot_by_sex(expression_ploidy_data, "log_KDM5D", "Y_ploidy")
create_gene_violin_plot_by_sex(expression_ploidy_data, "log_DDX3Y", "Y_ploidy")
create_gene_violin_plot_by_sex(expression_ploidy_data, "log_EIF1AY", "Y_ploidy")
create_gene_violin_plot_by_sex(expression_ploidy_data, "log_NLGN4Y", "Y_ploidy")
create_gene_violin_plot_by_sex(expression_ploidy_data, "log_RPS4Y1", "Y_ploidy")
create_gene_violin_plot_by_sex(expression_ploidy_data, "log_TMSB4Y", "Y_ploidy")
create_gene_violin_plot_by_sex(expression_ploidy_data, "log_USP9Y", "Y_ploidy")
create_gene_violin_plot_by_sex(expression_ploidy_data, "log_UTY", "Y_ploidy")
create_gene_violin_plot_by_sex(expression_ploidy_data, "log_ZFY", "Y_ploidy")
create_gene_violin_plot_by_sex(expression_ploidy_data, "log_XIST", "Y_ploidy")
```

```{r Plot X Ploidy na.omit}
create_gene_violin_plot_by_sex(expression_ploidy_data, "log_KDM5D", "X_ploidy")
create_gene_violin_plot_by_sex(expression_ploidy_data, "log_DDX3Y", "X_ploidy")
create_gene_violin_plot_by_sex(expression_ploidy_data, "log_EIF1AY", "X_ploidy")
create_gene_violin_plot_by_sex(expression_ploidy_data, "log_NLGN4Y", "X_ploidy")
create_gene_violin_plot_by_sex(expression_ploidy_data, "log_RPS4Y1", "X_ploidy")
create_gene_violin_plot_by_sex(expression_ploidy_data, "log_TMSB4Y", "X_ploidy")
create_gene_violin_plot_by_sex(expression_ploidy_data, "log_USP9Y", "X_ploidy")
create_gene_violin_plot_by_sex(expression_ploidy_data, "log_UTY", "X_ploidy")
create_gene_violin_plot_by_sex(expression_ploidy_data, "log_ZFY", "X_ploidy")
create_gene_violin_plot_by_sex(expression_ploidy_data, "log_XIST", "X_ploidy")
```

```{r}
create_gene_violin_plot_by_sex(male_data, "log_KDM5D", "Y_ploidy")
create_gene_violin_plot_by_sex(male_data, "log_DDX3Y", "Y_ploidy")
create_gene_violin_plot_by_sex(male_data, "log_EIF1AY", "Y_ploidy")
create_gene_violin_plot_by_sex(male_data, "log_NLGN4Y", "Y_ploidy")
create_gene_violin_plot_by_sex(male_data, "log_RPS4Y1", "Y_ploidy")
create_gene_violin_plot_by_sex(male_data, "log_TMSB4Y", "Y_ploidy")
create_gene_violin_plot_by_sex(male_data, "log_USP9Y", "Y_ploidy")
create_gene_violin_plot_by_sex(male_data, "log_UTY", "Y_ploidy")
create_gene_violin_plot_by_sex(male_data, "log_ZFY", "Y_ploidy")
create_gene_violin_plot_by_sex(male_data, "log_XIST", "Y_ploidy")

```

```{r}
create_gene_violin_plot_by_sex(male_data, "log_KDM5D", "X_ploidy")
create_gene_violin_plot_by_sex(male_data, "log_DDX3Y", "X_ploidy")
create_gene_violin_plot_by_sex(male_data, "log_EIF1AY", "X_ploidy")
create_gene_violin_plot_by_sex(male_data, "log_NLGN4Y", "X_ploidy")
create_gene_violin_plot_by_sex(male_data, "log_RPS4Y1", "X_ploidy")
create_gene_violin_plot_by_sex(male_data, "log_TMSB4Y", "X_ploidy")
create_gene_violin_plot_by_sex(male_data, "log_USP9Y", "X_ploidy")
create_gene_violin_plot_by_sex(male_data, "log_UTY", "X_ploidy")
create_gene_violin_plot_by_sex(male_data, "log_ZFY", "X_ploidy")
create_gene_violin_plot_by_sex(male_data, "log_XIST", "X_ploidy")

```

```{r}
create_gene_violin_plot_by_sex(female_data, "log_KDM5D", "Y_ploidy")
create_gene_violin_plot_by_sex(female_data, "log_DDX3Y", "Y_ploidy")
create_gene_violin_plot_by_sex(female_data, "log_EIF1AY", "Y_ploidy")
create_gene_violin_plot_by_sex(female_data, "log_NLGN4Y", "Y_ploidy")
create_gene_violin_plot_by_sex(female_data, "log_RPS4Y1", "Y_ploidy")
create_gene_violin_plot_by_sex(female_data, "log_TMSB4Y", "Y_ploidy")
create_gene_violin_plot_by_sex(female_data, "log_USP9Y", "Y_ploidy")
create_gene_violin_plot_by_sex(female_data, "log_UTY", "Y_ploidy")
create_gene_violin_plot_by_sex(female_data, "log_ZFY", "Y_ploidy")
create_gene_violin_plot_by_sex(female_data, "log_XIST", "Y_ploidy")

```

```{r}
create_gene_violin_plot_by_sex(female_data, "log_KDM5D", "X_ploidy")
create_gene_violin_plot_by_sex(female_data, "log_DDX3Y", "X_ploidy")
create_gene_violin_plot_by_sex(female_data, "log_EIF1AY", "X_ploidy")
create_gene_violin_plot_by_sex(female_data, "log_NLGN4Y", "X_ploidy")
create_gene_violin_plot_by_sex(female_data, "log_RPS4Y1", "X_ploidy")
create_gene_violin_plot_by_sex(female_data, "log_TMSB4Y", "X_ploidy")
create_gene_violin_plot_by_sex(female_data, "log_USP9Y", "X_ploidy")
create_gene_violin_plot_by_sex(female_data, "log_UTY", "X_ploidy")
create_gene_violin_plot_by_sex(female_data, "log_ZFY", "X_ploidy")
create_gene_violin_plot_by_sex(female_data, "log_XIST", "X_ploidy")

```

```{r}
# Run the function for each gene for Y_ploidy and save the plots
genes <- c("log_KDM5D", "log_DDX3Y", "log_EIF1AY", "log_NLGN4Y", 
           "log_RPS4Y1", "log_TMSB4Y", "log_USP9Y", "log_UTY", "log_ZFY", "log_XIST")

# Create and save plots for each gene
for (gene in genes) {
 all_data_plot <- create_gene_violin_plot_by_sex(expression_ploidy_data, gene, "Y_ploidy")
  
  # Save the plot to a PNG file
  ggsave(filename = file.path(output_directory, paste0(gene, "_Y_ploidy_sex_violin_plot.png")), plot = all_data_plot, width = 8, height = 6)
}

# Run the function for each gene for X_ploidy and save the plots
genes <- c("log_KDM5D", "log_DDX3Y", "log_EIF1AY", "log_NLGN4Y", 
           "log_RPS4Y1", "log_TMSB4Y", "log_USP9Y", "log_UTY", "log_ZFY", "log_XIST")

# Create and save plots for each gene
for (gene in genes) {
  all_data_plot <- create_gene_violin_plot_by_sex(expression_ploidy_data, gene, "X_ploidy")
  
  # Save the plot to a PNG file
  ggsave(filename = file.path("output_directory", paste0("All_", gene, "_Y_ploidy_violin_plot.png")), 
         plot = all_data_plot, width = 8, height = 6)
}


```

```{r}

# Create and save plots for males
for (gene in genes) {
  male_plot <- create_gene_violin_plot_by_sex(male_data, gene, "Y_ploidy")
  
  # Save the male plot to a PNG file
  ggsave(filename = file.path("output_directory", paste0("Male_", gene, "_Y_ploidy_violin_plot.png")), 
         plot = male_plot, width = 8, height = 6)
}

# Create and save plots for females
for (gene in genes) {
  female_plot <- create_gene_violin_plot_by_sex(female_data, gene, "Y_ploidy")
  
  # Save the female plot to a PNG file
  ggsave(filename = file.path("output_directory", paste0("Female_", gene, "_Y_ploidy_violin_plot.png")), 
         plot = female_plot, width = 8, height = 6)
}


```

```{r}

# Create and save plots for males
for (gene in genes) {
  male_plot <- create_gene_violin_plot_by_sex(male_data, gene, "X_ploidy")
  
  # Save the male plot to a PNG file
  ggsave(filename = file.path("output_directory", paste0("Male_", gene, "_X_ploidy_violin_plot.png")), 
         plot = male_plot, width = 8, height = 6)
}

# Create and save plots for females
for (gene in genes) {
  female_plot <- create_gene_violin_plot_by_sex(female_data, gene, "X_ploidy")
  
  # Save the female plot to a PNG file
  ggsave(filename = file.path("output_directory", paste0("Female_", gene, "_X_ploidy_violin_plot.png")), 
         plot = female_plot, width = 8, height = 6)
}


```

```{r}
# Function to create faceted violin plots for multiple gene expression variables by sex
create_faceted_violin_plot <- function(data, ploidy, remove_na = TRUE) {
  # Optionally filter out rows where ploidy is NA
  if (remove_na) {
    data <- data[!is.na(data[[ploidy]]), ]
  }

  # Filter out rows with "Miss-classification" in Biological_sex
  data <- data[data$Biological_sex != "Miss_classification", ]

  # List of genes to be plotted
  genes <- c("log_KDM5D", "log_DDX3Y", "log_EIF1AY", "log_NLGN4Y", "log_RPS4Y1", 
             "log_TMSB4Y", "log_USP9Y", "log_UTY", "log_ZFY", "log_XIST")

  # Convert the data to long format for faceting
  data_long <- tidyr::pivot_longer(data, cols = all_of(genes), names_to = "gene_expression", values_to = "expression_value")

  # Create spaced interaction for x-axis
  data_long$spaced_interaction <- as.factor(paste(data_long$Biological_sex, data_long[[ploidy]], " ", sep = ""))

  # Plot the data with faceting by gene
  ggplot(data_long, aes(x = spaced_interaction, y = expression_value, fill = Biological_sex)) + 
    geom_violin(width = 0.8, scale = "width") +  # Adjust width for better appearance
    labs(
      title = "Distribution of Gene Expression by Reported Sex and Ploidy",
      x = "Reported Sex and Ploidy Status",
      y = "Log-transformed Expression"
    ) +
    scale_fill_manual(values = c("Female" = "darkblue", "Male" = "lightblue"), name = "Reported Sex") + 
    theme_minimal() +
    scale_y_continuous(labels = function(x) ifelse(is.finite(x), round(x, 2), NA)) +
    scale_x_discrete(name = "Reported Sex and Ploidy Status") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    facet_wrap(~ gene_expression, scales = "free_y")  # Facet by gene expression
}

```

```{r}
# Create faceted violin plot for all data
all_data_plot <- create_faceted_violin_plot(expression_ploidy_data, "Y_ploidy")

all_data_plot

```
```{r}
# Create faceted violin plot for all data
all_data_Y_plot <- create_faceted_violin_plot(expression_ploidy_data, "Y_ploidy")
all_data_X_plot <- create_faceted_violin_plot(expression_ploidy_data, "X_ploidy")
all_data_Y_plot
all_data_X_plot

```

```{r}
# Create faceted violin plot for male data
male_data_Y_plot <- create_faceted_violin_plot(male_data, "Y_ploidy")
male_data_X_plot <- create_faceted_violin_plot(male_data, "X_ploidy")
male_data_Y_plot
male_data_X_plot

```

```{r}
# Create faceted violin plot for male data
female_data_Y_plot <- create_faceted_violin_plot(female_data, "Y_ploidy")
female_data_X_plot <- create_faceted_violin_plot(female_data, "X_ploidy")
female_data_Y_plot
female_data_X_plot

```


```{r}
# Save the plot as a PNG file
ggsave(filename = file.path("output_directory", "Faceted_Gene_Expression_Y_Ploidy_Violin_Plot.png"), 
       plot = all_data_Y_plot, width = 12, height = 8)

ggsave(filename = file.path("output_directory", "Faceted_Gene_Expression_X_Ploidy_Violin_Plot.png"), 
       plot = all_data_X_plot, width = 12, height = 8)

```

```{r}
# Save the plot as a PNG file
ggsave(filename = file.path("output_directory", "Male_Faceted_Gene_Expression_Y_Ploidy_Violin_Plot.png"), 
       plot = male_data_Y_plot, width = 12, height = 8)

ggsave(filename = file.path("output_directory", "Male_Faceted_Gene_Expression_X_Ploidy_Violin_Plot.png"), 
       plot = male_data_X_plot, width = 12, height = 8)

```

```{r}
# Save the plot as a PNG file
ggsave(filename = file.path("output_directory", "Female_Faceted_Gene_Expression_Y_Ploidy_Violin_Plot.png"), 
       plot = female_data_Y_plot, width = 12, height = 8)

ggsave(filename = file.path("output_directory", "Female_Faceted_Gene_Expression_X_Ploidy_Violin_Plot.png"), 
       plot = female_data_X_plot, width = 12, height = 8)

```
